---
title: "Major Diseases causing Deaths in Kenya"
author: "Sammy kariuki"
date: "March 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Introduction 
This is an exploration of dataset consisting number of deaths caused by major diseases.  
The data is available to the public here https://data.humdata.org/dataset.     

Before we start any data exploration and analysis lets lets load all the library we will need for our exploration.     
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
```
Now our environment is ready for us to dive into work.    
   
We start by creating a directory where we will store our data.   
```{r}
if (!file.exists("data")){
  dir.create("data")
}
```
Then move to the source of our dataset and get the link which we will use  to download the data into our working machine.    
```{r}
file_url <- "https://data.humdata.org/dataset/800a36ce-3dac-4266-9afc-6233ec2f3bd5/resource/9aa2b2eb-5a34-4fbf-bd62-e605dbaf3400/download/national-registered-deaths-by-major-causes-2010-2013.csv"  

if (!file.exists("./data/dat1.csv")){
  download.file(file_url, destfile = "data/dat1.csv") 
  downloadtime <- date()
}
```
Each time we download data from internet it is a good practice to record the time we download, this help you in future when you redownload the data and find it has being updated.   
```{r}
#print (downloadtime)
```
Lets read the data into R. 
 
```{r}
mydf1 <- read.csv("./data/dat1.csv")
```
Now lets explore our data.   
```{r}
class(mydf1)
dim(mydf1)
head(mydf1)
```
Now we can see.   
+ We have a dataframe as our R object   
+ We have five column with rows    
+ The first column is the main diseases while the rest are years and t
he number of deaths for each disease.   
    
    
##Tidy Data
Like families, tidy datasets are all alike but every messy dataset is messy in its own way. Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning). 
In **tidy data**   
1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.   
     
These principles of tidy data provide a standard way to organise data values within a dataset. A standard makes initial data cleaning easier because you don't need to start from scratch and reinvent the wheel every time.   
Back to our data we can clearly conclude that its quite messy thus we need to make it tidy.   
From the data we have three variables that is    
+ Cause   
+ Year   
+ Deaths   
This means that the data violate the rule that tidy data each variable form a column.   
Lets dive in to achieve this using the dplyr and tidyr package.    
We begin by using "tbl_df" in dplyr package that create a dataframe that ffers better checking and printing capabilities than traditional data frames.   
```{r}
mydf <- tbl_df(mydf1)
class(mydf)
mydf
```
Nice the new dataframe provide us with more info neatly.   
    
Further to avoid any confusion remove the former dataframe.  
```{r}
rm(mydf1)
```
Next we will utilise gather function tidyr package to clean the messy data.   
```{r}
tidy_mydf <- gather(mydf, Year, Deaths, -Cause)
dim(tidy_mydf)
tidy_mydf
```
Greate we now got three column this, data is tidy. But a closer look at Year column each year is proceeding by a letter x which is not very neat. Lets combine parse_number function in readr package and mutate in dplyr.   
```{r}
tidy_mydf <- mutate(tidy_mydf, Year = parse_number(Year))
tidy_mydf
```
Finally we have the Tidy data ready for further analysis lets write and save in a file for future use.
```{r}
write.csv(tidy_mydf, file = "data/tidy_dat1.csv")
```
## Including Plots for visualization
For a visual of the data let utilize ggplot2 package to embed a simple plot.   

```{r}
ggplot(data= tidy_mydf, mapping = aes(x=Year, y =Deaths, colors(Cause)))+
  geom_point(aes(colour = Cause )) +
  geom_line(aes(colour = Cause))
```

